Supplementary Information

### 2020-08-28 doc/manuscript/MEE_manuscript.Rmd
fixing up for coauthor review

@@

2. Here, we develop a new method that allows to keep the raw tree ring measurements and individual tree's metrics to simultaneously model non-linear effects of objectively determined principle climate drivers and e.g. tree diameter. 
Specifically, we first identify the most important climate drivers and their appropriate time window of influence using the climwin R package.
We then include these in generalized least squares models to model tree rings (or derived measurements) while accounting for the temporal autocorrelation inherent to each individual tree's growth.
We apply this method to tree-ring data from **#** species at ten globally distributed sites spanning a wide range of forest types.

3. Our analysis identified similar climate drivers to those obtained via traditional methods, but revealed that non-linear responses to climate variables were common. Radial growth increments, basal area increments, and biomass increments all varied non-linearly with tree diameter. [*something about interactions between climate sensitivity and diameter*]

@@

@@

**Traditional methods do not characterize the effect of tree size or its potential interactions with climate variables.** 
Tree size (most commonly diameter breast height, $DBH$) is among the most important variables affecting tree growth rate [@muller-landau_testing_2006; @foster_predicting_2016; REFS]. 
Radial growth increments (*i.e.*, tree-ring widths; $\Delta r$) may increase or decrease with tree size, often in a non-linear manner.
Following a "juvenile growth phase", which is typically removed in traditional dendrological analyses, $\Delta r$ may decline, particularly in open-grown conifers (??; DENDRO_REFS).
In contrast, in mesic closed-canopy forests, $\Delta r$ typically increases with tree diameter at breast height [$DBH$; @muller-landau_testing_2006; @anderson-teixeira_size-related_2015; REFS, DENDRO_REFS].
For dendrological studies aimed at deciphering climate signals, tree size is not typically a variable of interest, and its influence is removed through detrending (DENDRO_REFS).
While suitable for identifying climate signals (DENDRO_REFS), this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity. 
Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016]  and scaled to characterize the climate sensitivity of $ANPP_{stem}$ based on the scaling of $\Delta r$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
For example, larger trees tend to exhibit greater growth reductions in drought years [@bennett_larger_2015 ; @mcgregor_tree_nodate].
Thus, to use tree-rings to predict $ANPP_{stem}$ for a forest where not every tree was cored, we need models that include tree size.

@@

**Table 1 | Sites included in this analysis**
![ ](tables_figures/sites_table.png)
```{r eval=FALSE, echo=FALSE}
# refs in this table need to go in the nocite metadata field at the top of the document
```

All tree cores were measured and cross-dated by the original researchers using standard dendrochronological practices. 
The full record for all cores was retained for analysis, with the exception of records that had to be excluded due to insufficient sample size or anomalous growth patterns.
Specifically, we excluded
(1) species with <7 cores,
(2) cores with <30 years of record,
(3) parts of records before or after (whichever was shorter) any year where $\Delta r$ exceeded the mean plus five times the standard deviation of $\Delta r$ for the entire core, along with a 5-year buffer in the opposite direction, *(this part needs rewording)*
and (4) the final 20 years prior to death of trees cored dead. 
The final criteria was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017] ; REF on climate sensitivity).
From analyses including $DBH$ (see below), we further excluded
(1) trees for which we lacked data required to reconstruct $DBH$,
(2) trees for which there was a significant inconsistency between measured $DBH$ and the sum of $\Delta r$'s across the core (*Appendix S1*),
(3) parts of records where reconstructed $DBH$ (see below) was represented by <3 conspecific trees.

@@

Once $DBH$ had been reconstructed, we used biomass allometries to estimate the corresponding aboveground biomass and diameter to area equation to get the corresponding basal area. We then calculated aboveground biomass growth increments ($\Delta AGB$) as [$AGB_{y+1}-AGB_y$] and basal area increment ($BAI$) as [$BAI_{y+1}-BAI_y$].
For temperate sites, biomass allometries were sourced from allo-db (DETAILS; Gonzalez-Akre et al. in prep).
For tropical sites, biomass allometries were sourced from the BIOMASS package (REF).

Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020]. 
Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); potential evapotranspiration ($PET$); precipitation ($PPT$); and precipitation day frequency ($PDF$). 
For BCI, we calculated $PPT$ and $PDF$ from daily precipitation readings made on BCI starting in 1929 [@paton_barro_2019].
All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0-alpha [@anderson-teixeira_forestgeoclimate_2020].

>>>>

*NOTE TO COAUTHORS: I'd appreciate feedback on what you consider to be the most important results and conclusions to highlight in the abstract. I think we have many interesting findings with important implications.*

@@

2. Here, we develop a new method to simultaneously model non-linear effects of objectively determined principle climate drivers and individual tree metrics (e.g., diameter). 
Specifically, we first identify the most important climate drivers and their appropriate time window of influence using the climwin R package.
We then include these in generalized least squares models to model tree growth while accounting for the temporal autocorrelation inherent to each individual tree's growth.
We apply this method to tree-ring data from *~40* species at ten globally distributed sites spanning a wide range of forest types.

3. Our analysis identified similar climate drivers to those obtained via traditional methods, but revealed that non-linear responses to climate variables were common. Radial growth increments, basal area increments, and biomass increments all varied non-linearly with tree diameter. The relative importance of tree diameter versus climate varied across sites, and interactions between diameter and climate variables were prevalent (*~50%* of cases tested).

@@

*NOTE TO COAUTHORS: Intro is still pretty rough, so please don't worry about word-smithing. However, I'd appreciate input on the content, and important references (especially "DENDRO REFS", as most of you know that literature better than I do).*

@@

**Traditional methods do not characterize the effect of tree size or its potential interactions with climate variables.** 
Tree size (most commonly diameter breast height, $DBH$) is among the most important variables affecting tree growth rate [@muller-landau_testing_2006; @foster_predicting_2016; REFS]. 
Radial growth increments (*i.e.*, tree-ring widths; $\Delta r$) may increase or decrease with tree size, often in a non-linear manner.
Following a "juvenile growth phase", which is typically removed in traditional dendrological analyses, $\Delta r$ may decline, particularly in open-grown conifers (??; DENDRO_REFS).
In contrast, in mesic closed-canopy forests, $\Delta r$ typically increases with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; REFS, DENDRO_REFS].
For dendrological studies aimed at deciphering climate signals, tree size is not typically a variable of interest, and its influence is removed through detrending (DENDRO_REFS).
While suitable for identifying climate signals (DENDRO_REFS), this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity. 
Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016]  and scaled to characterize the climate sensitivity of $ANPP_{stem}$ based on the scaling of $\Delta r$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
For example, larger trees tend to exhibit greater growth reductions in drought years [@bennett_larger_2015 ; @mcgregor_tree_nodate].
Thus, to use tree-rings to predict $ANPP_{stem}$ for a forest where not every tree was cored, we need models that include tree size.

@@

**Table 1 | Sites included in this analysis**. Here and throughout, sites are ordered by latitude. Additional site information is provided in Table S1, and species and their sample sizes are detailed in Tables S2-S3. *Climate will be filled in based on CRU data.*
![ ](tables_figures/sites_table.png)
```{r eval=FALSE, echo=FALSE}
# refs in this table need to go in the nocite metadata field at the top of the document
```

All tree cores were measured and cross-dated by the original researchers using standard dendrochronological practices. 
The full record for all cores was retained for analysis, with the exception of records that had to be excluded due to insufficient sample size or anomalous growth patterns.
Specifically, we excluded
(1) species with <7 cores,
(2) cores with <30 years of record,
(3) parts of records before or after (whichever was shorter) any year where $\Delta r$ exceeded the mean plus five times the standard deviation of $\Delta r$ for the entire core, along with a 5-year buffer in the opposite direction *(this part needs rewording)*,
and (4) the final 20 years prior to death of trees cored dead. 
The final criteria was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017] ; REF on climate sensitivity).
From analyses including $DBH$ (see below), we further excluded
(1) trees for which we lacked data required to reconstruct $DBH$,
(2) trees for which there was a significant inconsistency between measured $DBH$ and the sum of $\Delta r$'s across the core (*Appendix S1*),
(3) parts of records where reconstructed $DBH$ (see below) was represented by <3 conspecific trees.
In total, this resulted in inclusion of *#* cores, *#* of which could be included in analyses with $DBH$ (Table S3).

@@

Once $DBH$ had been reconstructed, we used biomass allometries to estimate the corresponding aboveground biomass and diameter to area equation to get the corresponding basal area. We then calculated aboveground biomass growth increments ($\Delta AGB$) as [$AGB_{y+1}-AGB_y$] and basal area increment ($BAI$) as [$BAI_{y+1}-BAI_y$].
Biomass allometries for temperate and tropical sites were calculated using the *allo-db* (Gonzalez-Akre et al. in prep) and *biomass* [@rejoumechain_biomass_2017] R packages, respectively.

Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020]. 
*Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$ / currently "TMN", $T_{max}$ / currently "TMX", $T_{mean}$/ / currently "TMP", respectively); potential evapotranspiration ($PET$); precipitation ($PPT$/ currently "PRE"); and precipitation day frequency ($PDF$/ currently "WET").* 
For BCI, we calculated $PPT$ and $PDF$ from daily precipitation readings made on BCI starting in 1929 [@paton_barro_2019].
All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0-alpha [@anderson-teixeira_forestgeoclimate_2020].
